---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">",
  fig.path = "man/figures/README-"
)
```

```{r library, echo = FALSE}
library(packageRank)
```

## packageRank: compute and visualize package download counts and rank percentiles

['packageRank'](https://CRAN.R-project.org/package=packageRank) is an R package that helps put package download counts into context. It does so via two functions, `cranDownloads()` and `packageRank()`, and one set of filters that remove "invalid" entries from the download logs. I cover these topics in three parts. A fourth part covers package related issues.

* [Part I Package Download Counts](#i---computing-package-download-counts) describes how `cranDownloads()` extends the functionality of [`cranlogs::cran_downloads()`](https://r-hub.github.io/cranlogs/) by adding a more user-friendly interface and providing generic R `plot()` methods to make visualization easy.
* [Part II Package Download Rank Percentiles](#ii---computing-package-download-rank-percentiles) describes how `packageRank()` uses rank percentiles, a nonparametric statistic that tells you the percentage of observations (i.e., packages) with fewer counts (i.e., downloads), to  help you see how your package is doing relative to _all_ other [CRAN](https://CRAN.R-project.org/) packages.
* [Part III Package Download Filters](#iii---filtering-package-download-counts) describes how filters are used to reduce the presence of software and behavioral artifacts that would otherwise inflate package download counts.
* [Part IV Notes](#iv---notes) describes some technical issues, including the use of memoization, time zone issues, and internet connection time out problems.

[‘packageRank'](https://CRAN.R-project.org/package=packageRank) requires an active internet connection, and relies on the [‘cranlogs'](https://CRAN.R-project.org/package=cranlogs) package and [RStudio's download logs](http://cran-logs.rstudio.com/). The latter records traffic to the “0-Cloud” mirror at cloud.R-project.org, which is “currently sponsored by RStudio” and was formerly RStudio's CRAN mirror.

Note that logs for the previous day are generally posted by 17:00 UTC. Updated results for functions that rely on [‘cranlogs'](https://CRAN.R-project.org/package=cranlogs) are typically available soon thereafter.

### getting started

To install ['packageRank'](https://cran.r-project.org/package=packageRank) from CRAN:

```{r cran_install, eval = FALSE}
install.packages("packageRank")
```

To install the development version from GitHub:

```{r gh_install, eval = FALSE}
# You may need to first install 'remotes' via install.packages("remotes").
remotes::install_github("lindbrook/packageRank", build_vignettes = TRUE)
```

### I - computing package download counts

`cranDownloads()` uses all the same arguments as `cranlogs::cran_downloads()`:

```{r cran_downloads, eval = FALSE}
cranlogs::cran_downloads(packages = "HistData")
```
```{r cran_downloads_code, echo = FALSE}
cranlogs::cran_downloads(packages = "HistData", from = "2020-05-01", to = "2020-05-01")
```

The only difference is that `cranDownloads()` adds four features:

#### i) "spell check" for package names

```{r spell_check_fail, eval = FALSE}
cranDownloads(packages = "GGplot2")
```
```
## Error in cranDownloads(packages = "GGplot2") :
##   GGplot2: misspelled or not on CRAN.
```

<br/>
```{r spell_check_pass, eval = FALSE}
cranDownloads(packages = "ggplot2")
```
```{r spell_check_pass_code, echo = FALSE}
cranDownloads(packages = "ggplot2", from = "2020-05-01", to = "2020-05-01")
```

<br/>
This also works for inactive or "retired" packages in the [Archive](https://CRAN.R-project.org/src/contrib/Archive/):

```{r check_archive_fail, eval = FALSE}
cranDownloads(packages = "vr")
```
```
## Error in cranDownloads(packages = "vr") :
##  vr: misspelled or not on CRAN/Archive.
```

<br/>
```{r check_archive_pass, eval = FALSE}
cranDownloads(packages = "VR")
```
```{r check_archive_pass_code, echo = FALSE}
cranDownloads(packages = "VR", from = "2020-05-01", to = "2020-05-01")
```

<br/>

#### ii) two additional date formats

With `cranlogs::cran_downloads()`, you specify a time frame using the `from` and `to` arguments. The downside of this is that you _must_ use the "yyyy-mm-dd" date format. For convenience's sake, `cranDownloads()` also allows you to use "yyyy-mm" or "yyyy" (yyyy also works).

##### "yyyy-mm"
Let's say you want the download counts for ['HistData'](https://CRAN.R-project.org/package=HistData) for the month of February 2020. With `cranlogs::cran_downloads()`, you'd have to type out the whole date and remember that 2020 was a leap year:

```{r yyyy-mm_1, eval = FALSE}
cranlogs::cran_downloads(packages = "HistData", from = "2020-02-01",
  to = "2020-02-29")
```

<br/>
With `cranDownloads()`, you can just specify the year and month:

```{r yyyy-mm_2, eval = FALSE}
cranDownloads(packages = "HistData", from = "2020-02", to = "2020-02")
```

##### "yyyy" or yyyy
Let's say you want the year-to-date download counts for ['rstan'](https://CRAN.R-project.org/package=rstan). With `cranlogs::cran_downloads()`, you'd type something like:

```{r yyyy_1, eval = FALSE}
cranlogs::cran_downloads(packages = "rstan", from = "2021-01-01",
  to = Sys.Date() - 1)
```

<br/>
With `cranDownloads()`, you can use:

```{r yyyy_2, eval = FALSE}
cranDownloads(packages = "rstan", from = "2021")
```

or

```{r yyyy_3, eval = FALSE}
cranDownloads(packages = "rstan", from = 2021)
```

<br/>

#### iii) check date validity

```{r check_date, eval = FALSE}
cranDownloads(packages = "HistData", from = "2019-01-15",
  to = "2019-01-35")
```
```
## Error in resolveDate(to, type = "to") : Not a valid date.
```

#### iv) cumulative count

```{r cranDownloads_cumulative, eval = FALSE}
cranDownloads(packages = "HistData", when = "last-week")
```

```{r cranDownloads_cumulative_code, echo = FALSE}
cranDownloads(packages = "HistData", from = "2020-05-01", to = "2020-05-07")
```


<br/>

### visualizing package download counts

`cranDownloads()` makes visualizing package downloads easy. Just use `plot()`:

```{r cranDownloads_viz1, fig.align = "center"}
plot(cranDownloads(packages = "HistData", from = "2019", to = "2019"))
```

If you pass a vector of package names for a single day, `plot()` will return a dotchart:

```{r cranDownloads_viz2a, fig.align = "center"}
plot(cranDownloads(packages = c("ggplot2", "data.table", "Rcpp"),
  from = "2020-03-01", to = "2020-03-01"))
```

If you pass a vector of package names for multiple days, `plot()` uses `ggplot2` facets:

```{r cranDownloads_viz2, fig.align = "center"}
plot(cranDownloads(packages = c("ggplot2", "data.table", "Rcpp"),
  from = "2020", to = "2020-03-20"))
```
<br/>

If you want to plot those data in a single frame, set `multi.plot = TRUE`:

```{r cranDownloads_viz3}
plot(cranDownloads(packages = c("ggplot2", "data.table", "Rcpp"),
  from = "2020", to = "2020-03-20"), multi.plot = TRUE)
```

<br/>
If you want plot those data in separate plots on the same scale, use `graphics = "base"` and you'll be prompted for each plot:

```{r cranDownloads_viz4, eval = FALSE}
plot(cranDownloads(packages = c("ggplot2", "data.table", "Rcpp"),
  from = "2020", to = "2020-03-20"), graphics = "base")
```

If you want do the above on separate independent scales, set `same.xy = FALSE`:

```{r cranDownloads_viz5, eval = FALSE}
plot(cranDownloads(packages = c("ggplot2", "data.table", "Rcpp"),
  from = "2020", to = "2020-03-20"), graphics = "base", same.xy = FALSE)
```

#### `packages = NULL`

`cranlogs::cran_download(packages = NULL)` computes the total number of package downloads from CRAN. You can plot these data by using:

```{r null_packages}
plot(cranDownloads(from = 2019, to = 2019))
```

#### `packages = "R"`

`cranlogs::cran_download(packages = "R")` computes the total number of downloads of the R application (note that you can only use "R" or a vector of packages names, not both!). You can plot these data by using:

```{r r_downloads}
plot(cranDownloads(packages = "R", from = 2019, to = 2019))
```

#### smoothers and confidence intervals

To add a lowess smoother to your plot, use `smooth = TRUE`:

```{r lowess}
plot(cranDownloads(packages = "rstan", from = "2019", to = "2019"),
  smooth = TRUE)
```

With graphs that use 'ggplot2', `se = TRUE` will add confidence intervals:

```{r ci}
plot(cranDownloads(packages = c("HistData", "rnaturalearth", "Zelig"),
  from = "2020", to = "2020-03-20"), smooth = TRUE, se = TRUE)
```

#### package and R release dates

To annotate a graph with a package's release dates:

```{r pkg_release_date}
plot(cranDownloads(packages = "rstan", from = "2019", to = "2019"),
  package.version = TRUE)
```

To annotate a graph with R release dates:

```{r r_release_date}
plot(cranDownloads(packages = "rstan", from = "2019", to = "2019"),
  r.version = TRUE)
```

#### plot growth curves (cumulative download counts)

To plot growth curves, set `statistic = "cumulative"`:

```{r cranDownloads_growth_curves}
plot(cranDownloads(packages = c("ggplot2", "data.table", "Rcpp"),
  from = "2020", to = "2020-03-20"), statistic = "cumulative",
  multi.plot = TRUE, points = FALSE)
```

#### population plot

To visualize a package's downloads relative to "all" other packages over time:

```{r pop_plot, eval = FALSE}
plot(cranDownloads(packages = "HistData", from = "2020", to = "2020-03-20"),
  population.plot = TRUE)
```

```{r pop_plot_code, echo = FALSE}
plot(cranDownloads(packages = "HistData", from = "2020", to = "2020-03-20"),
  population.plot = TRUE, population.seed = 1)
```

This longitudinal view of package downloads plots the date (x-axis) against the logarithm of a package's downloads (y-axis). In the background, the same variable are plotted (in gray) using a stratified random sample of packages: within each 5% interval of rank percentiles (e.g., 0 to 5, 5 to 10, 95 to 100, etc.), a random sample of 5% of packages is selected and tracked. This graphically approximates the "typical" pattern of downloads on CRAN for the selected time period.

### II - computing package download rank percentiles

After looking at nominal download counts for a while, the "compared to what?" question comes to mind. For instance, consider the data for the first week of March 2020:

```{r motivation, eval = FALSE}
plot(cranDownloads(packages = "cholera", from = "2020-03-01",
  to = "2020-03-07"))
```

```{r motivation_code, echo = FALSE, fig.align = "center"}
par(mar = c(5, 4, 4, 4))
plot(cranDownloads(packages = "cholera", from = "2020-03-01",
  to = "2020-03-07"))
par(mar = c(5, 4, 4, 2))
```

Do Wednesday and Saturday reflect surges of interest in the package or surges of traffic to [CRAN](https://CRAN.R-project.org/)? To put it differently, how can we know if a given download count is typical or unusual? One way to answer these questions is to locate your package in the frequency distribution of download counts.

Below are the distributions of logarithm of download counts for Wednesday and Saturday. The location of a vertical segment along the x-axis represents a download count and the height of a segment represents that download count's frequency. The location of ['cholera'](https://CRAN.R-project.org/package=cholera) in the distribution is highlighted in red.

```{r packageDistribution, echo = FALSE}
plot_package_distribution <- function(dat, xlim, ylim) {
  freq.dist <- dat$freq.dist
  freqtab <- dat$freqtab
  plot(freq.dist$count, freq.dist$frequency, type = "h", log = "x",
    xlab = "Downloads", ylab = "Frequency", xlim = xlim, ylim = ylim)
  axis(3, at = freqtab[1], cex.axis = 0.8, padj = 0.9, col.axis = "dodgerblue",
    col.ticks = "dodgerblue", labels = paste(names(freqtab[1]), "=",
    format(freqtab[1], big.mark = ",")))
  abline(v = freqtab[1], col = "dodgerblue", lty = "dotted")

  if (!is.null(dat$package)) {
    pkg.ct <- freqtab[names(freqtab) == dat$package]
    pkg.bin <- freqtab[freqtab == pkg.ct]
    axis(3, at = pkg.ct, labels = format(pkg.ct, big.mark = ","),
      cex.axis = 0.8, padj = 0.9, col.axis = "red", col.ticks = "red")
    abline(v = pkg.ct, col = grDevices::adjustcolor("red", alpha.f = 0.5))
    day <- weekdays(as.Date(dat$date), abbreviate = TRUE)
    title(paste0(dat$package, " @ ", dat$date, " (", day, ")"))
  } else title(paste("Distribution of Package Download Counts:", dat$date))
}

distn.data <- lapply(c("2020-03-04", "2020-03-07"), function(date) packageDistribution(package = "cholera", date = date))

xlim <- range(lapply(distn.data, function(x) x$freq.dist$count))
ylim <- range(lapply(distn.data, function(x) x$freq.dist$frequency))
```

```{r packageDistribution_wed, eval = FALSE}
plot(packageDistribution(package = "cholera", date = "2020-03-04"))
```
```{r packageDistribution_wed_code, echo = FALSE, fig.align = "center"}
plot_package_distribution(distn.data[[1]], xlim, ylim)
```

```{r packageDistribution_sat, eval = FALSE}
plot(packageDistribution(package = "cholera", date = "2020-03-07"))
```
```{r packageDistribution_sat_code, echo = FALSE, fig.align = "center"}
plot_package_distribution(distn.data[[2]], xlim, ylim)
```

While these plots give us a better picture of where ['cholera'](https://CRAN.R-project.org/package=cholera) is located, comparisons between Wednesday and Saturday are impressionistic at best: all we can confidently say is that the download counts for both days were greater than the mode.

To facilitate interpretation and comparison, I use the _rank percentile_ of a download count in place of the nominal download count. This nonparametric statistic tells you the percentage of packages with fewer downloads. In other words, it gives you the location of your package relative to the locations of all other packages. More importantly, by rescaling download counts to lie on the bounded interval between 0 and 100, rank percentiles make it easier to compare packages within and across distributions.

For example, we can compare Wednesday ("2020-03-04") to Saturday ("2020-03-07"):

```{r packageRank1}
packageRank(package = "cholera", date = "2020-03-04")
```

On Wednesday, we can see that ['cholera'](https://CRAN.R-project.org/package=cholera) had 38 downloads, came in 5,556th place out of 18,038 observed packages, and earned a spot in the 68th percentile.

```{r packageRank2}
packageRank(package = "cholera", date = "2020-03-07")
```

On Saturday, we can see that ['cholera'](https://CRAN.R-project.org/package=cholera) had 29 downloads, came in 3,061st place out of 15,950 observed packages, and earned a spot in the 80th percentile.

So contrary to what the nominal counts tell us, one could say that the interest in ['cholera'](https://CRAN.R-project.org/package=cholera) was actually greater on Saturday than on Wednesday.

#### computing rank percentile

To compute rank percentiles, I do the following. For each package, I tabulate the number of downloads and then compute the percentage of packages with fewer downloads. Here are the details using ['cholera'](https://CRAN.R-project.org/package=cholera) from Wednesday as an example:

```{r percentile}
pkg.rank <- packageRank(packages = "cholera", date = "2020-03-04")

downloads <- pkg.rank$freqtab

round(100 * mean(downloads < downloads["cholera"]), 1)
```

To put it differently:

```{r percentile2}
(pkgs.with.fewer.downloads <- sum(downloads < downloads["cholera"]))

(tot.pkgs <- length(downloads))

round(100 * pkgs.with.fewer.downloads / tot.pkgs, 1)
```

#### nominal ranks

In the example above, 38 downloads puts 'cholera' in 5,556th place among 18,038 observed packages. This rank is "nominal" because it's possible that multiple packages can have the same number of downloads. As a result, a package's nominal rank (but not its rank percentile) can be affected by its name. This is because packages with the same number of downloads are sorted in alphabetical order. Thus, 'cholera' benefits from the fact that it is 31st in the list of 263 packages with 38 downloads:

```{r nominal}
pkg.rank <- packageRank(packages = "cholera", date = "2020-03-04")
downloads <- pkg.rank$freqtab

which(names(downloads[downloads == 38]) == "cholera")
length(downloads[downloads == 38])
```

### visualizing package download rank percentiles

To visualize `packageRank()`, use `plot()`.

```{r packageRank_plot_wed, eval = FALSE}
plot(packageRank(packages = "cholera", date = "2020-03-04"))
```

```{r packageRank_data, echo = FALSE}
dat <- lapply(c("2020-03-04", "2020-03-07"), function(x) {
  packageRank("cholera", date = x)
})

freqtab1 <- dat[[1]]$freqtab
freqtab2 <- dat[[2]]$freqtab
xlim <- range(seq_along(freqtab1), seq_along(freqtab2))
ylim <- range(c(freqtab1), c(freqtab2))
```

```{r packageRank_plot_code_wed, echo = FALSE, fig.align = "center"}

freqtab <- dat[[1]]$freqtab
package.data <- dat[[1]]$package.data
pkg <- dat[[1]]$packages
date <- dat[[1]]$date
y.max <- freqtab[1]
q <- stats::quantile(freqtab)[2:4]

iqr <- vapply(c("75%", "50%", "25%"), function(id) {
  dat <- which(freqtab > q[[id]])
  dat[length(dat)]
}, numeric(1L))

plot(seq_along(freqtab), c(freqtab), type = "l", xlab = "Rank",
  ylab = "log10(Count)", log = "y", xlim = xlim, ylim = ylim)
abline(v = iqr, col = "black", lty = "dotted")
iqr.labels <- c("75th", "50th", "25th")
invisible(lapply(seq_along(iqr), function(i) {
  text(iqr[[i]], y.max / 2, labels = iqr.labels[i], cex = 0.75)
}))
abline(v = which(names(freqtab) == pkg), col = "red")
abline(h = freqtab[pkg], col = "red")
pct <- package.data[package.data$package == pkg, "percentile"]
pct.label <- paste0(round(pct, 2), "%")
axis(3, at = which(names(freqtab) == pkg), padj = 0.9, col.axis = "red",
  col.ticks = "red", labels = pct.label, cex.axis = 0.8)
axis(4, at = freqtab[pkg], col.axis = "red", col.ticks = "red",
  cex.axis = 0.8, labels = format(freqtab[pkg], big.mark = ","))
points(which(names(freqtab) == pkg), freqtab[pkg], col = "red")
points(which(names(freqtab) == names(freqtab[1])), y.max,
  col = "dodgerblue")
text(which(names(freqtab) == names(freqtab[1])), y.max, pos = 4,
  labels = paste(names(freqtab[1]), "=", format(freqtab[1],
  big.mark = ",")), cex = 0.8, col = "dodgerblue")
text(max(xlim), max(ylim),
  labels = paste("Tot = ", format(sum(freqtab), big.mark = ",")), cex = 0.8,
  col = "dodgerblue", pos = 2)
day <- weekdays(as.Date(date), abbreviate = TRUE)
title(main = paste0(pkg, " @ ", date, " (", day, ")"))
```

<br/>

```{r packageRank_plot_sat, eval = FALSE}
plot(packageRank(packages = "cholera", date = "2020-03-07"))
```

```{r packageRank_plot_code_sat, echo = FALSE, fig.align = "center"}
freqtab <- dat[[2]]$freqtab
package.data <- dat[[2]]$package.data
pkg <- dat[[2]]$packages
date <- dat[[2]]$date
y.max <- freqtab[1]
q <- stats::quantile(freqtab)[2:4]

iqr <- vapply(c("75%", "50%", "25%"), function(id) {
  dat <- which(freqtab > q[[id]])
  dat[length(dat)]
}, numeric(1L))

plot(seq_along(freqtab), c(freqtab), type = "l", xlab = "Rank",
  ylab = "log10(Count)", log = "y", xlim = xlim, ylim = ylim)
abline(v = iqr, col = "black", lty = "dotted")
iqr.labels <- c("75th", "50th", "25th")
invisible(lapply(seq_along(iqr), function(i) {
  text(iqr[[i]], y.max / 2, labels = iqr.labels[i], cex = 0.75)
}))
abline(v = which(names(freqtab) == pkg), col = "red")
abline(h = freqtab[pkg], col = "red")
pct <- package.data[package.data$package == pkg, "percentile"]
pct.label <- paste0(round(pct, 2), "%")
axis(3, at = which(names(freqtab) == pkg), padj = 0.9, col.axis = "red",
  col.ticks = "red", labels = pct.label, cex.axis = 0.8)
axis(4, at = freqtab[pkg], col.axis = "red", col.ticks = "red",
  cex.axis = 0.8, labels = format(freqtab[pkg], big.mark = ","))
points(which(names(freqtab) == pkg), freqtab[pkg], col = "red")
points(which(names(freqtab) == names(freqtab[1])), y.max,
  col = "dodgerblue")
text(which(names(freqtab) == names(freqtab[1])), y.max, pos = 4,
  labels = paste(names(freqtab[1]), "=", format(freqtab[1],
  big.mark = ",")), cex = 0.8, col = "dodgerblue")
text(max(xlim), max(ylim),
  labels = paste("Tot = ", format(sum(freqtab), big.mark = ",")), cex = 0.8,
  col = "dodgerblue", pos = 2)
day <- weekdays(as.Date(date), abbreviate = TRUE)
title(main = paste0(pkg, " @ ", date, " (", day, ")"))
```

These graphs, customized to be on the same scale, plot the _rank order_ of packages' download counts (x-axis) against the logarithm of those counts (y-axis). It then highlights a package's position in the distribution along with its rank percentile and download count (in red). In the background, the 75th, 50th and 25th percentiles are plotted as dotted vertical lines. The package with the most downloads, ['magrittr'](https://CRAN.R-project.org/package=magrittr) in both cases, is at top left (in blue). The total number of downloads is at the top right (in blue).

### III - filtering package download counts

Package downloads are computed by counting the number of log entries for each package. While straightforward, this approach can run into problems. Putting aside questions surrounding package dependencies, here I'm focussing on what I believe are two sets of "invalid" log entries. The first, a software artifact, stems from entries that are smaller, often orders of magnitude smaller, than a package's actual binary or source file size. Here, the problem is that the nominal count wrongly credits these downloads. The second, a behavioral artifact, emerges from efforts to download all of the packages on [CRAN](https://cran.r-project.org/). Here, the problem is that you get an inflated sense of interest in your package.

An early but detailed analysis and discussion of both inflations is available as part of this [R-hub blog post](https://blog.r-hub.io/2020/05/11/packagerank-intro/#inflationary-bias-of-download-counts).

#### software artifacts

When looking at package download logs, the first thing you'll notice are wrongly sized log entries. They come in two sizes: "small" and "medium". The "small" entries are approximately 500 bytes. The "medium" entries are variable in size. They fall anywhere between a "small" and a full download (i.e., "small" <= "medium" <= full download). "Small" entries manifest themselves as standalone entries, as part of pair with a full download, or as part of a triplet with a "medium" and a full download. "Medium" entries manifest themselves as standalone entries, or as part of the aforementioned triplet.

The example below illustrates a triplet:

```{r triplet}
packageLog(date = "2020-07-01")[4:6, -(4:6)]
```

The "medium" entry is the first observation (99,622 bytes). The observed full download is the second entry (4,161,948 bytes). The "small" entry is the last observation (536 bytes). Incidentally, what makes a triplet a triplet (or a pair a pair) is that all members have, at a minimum, identical or adjacent time stamps.

To deal with the inflationary effect of "small" entries, I filter out observations smaller than 1,000 bytes (the smallest package appears to be ['source.gist'](https://cran.r-project.org/package=source.gist), which weighs in at 1,200 bytes). "Medium" entries are harder to handle. I remove them using either a triplet-specific filter or a filter that looks up a package's size.

#### behavioral artifacts

While wrongly sized entries are fairly easy to spot, seeing other types of "invalid" entries can sometimes require a change of perspective. What I have in mind here are downloads that are a consequence of efforts to download all of [CRAN](https://cran.r-project.org/): _all_ packages including _all_ past versions. For details and evidence see the [R-hub blog post](https://blog.r-hub.io/2020/05/11/packagerank-intro/#inflationary-bias-of-download-counts) mentioned above (I believe this excludes mirroring activity via `rsync`).

Consider the example below:

```{r, sequence_ex}
packageLog(packages = "cholera", date = "2020-07-31")[8:14, -(4:6)]
```

Here, we see that seven different versions of the package were downloaded in a sequential bloc. A little digging show that these seven versions represent _all_ prior versions of 'cholera':

```{r, cholera_history}
packageHistory(package = "cholera")
```

While there are legitimate reasons for downloading past versions (e.g., research, container-based software distribution, etc.), examples like the above are "fingerprints" of efforts to download [CRAN](https://cran.r-project.org/). The upshot here is that when your package is downloaded as part of such efforts, that download is more a reflection of an interest in [CRAN](https://cran.r-project.org/) as collection of packages than an interest in your package _per se_. And since one of the uses of counting package downloads is to estimate interest in your package, it may be useful to exclude such entries.

To do so, I try to filter out these entries in two ways. The first identifies IP addresses that download "too many" packages and then filters out "campaigns", large blocs of downloads that occur in (nearly) alphabetical order. The second looks for campaigns not associated with "greedy" IP addresses and filters out sequences of past versions downloaded in a narrowly defined time window.

#### example usage

To get an idea of how inflated your package's download count may be, use `filteredDownloads()`. Below are the results for 'cholera' for 31 July 2020.

```{r, filteredDownloads}
filteredDownloads(package = "cholera", date = "2020-07-31")
```

While there were 14 nominal downloads, applying all the filters reduced the number of downloads to 5, an inflation of 180%.

Note that the filters are computationally demanding. Excluding the time it takes to download the log file, the filters in the above example take approximate 75 seconds to run using parallelized code (currently only available on macOS and Unix) on a 3.1 GHz Dual-Core Intel Core i5 processor.

Currently, there are 5 different functions. They are controlled by the following function arguments (listed in order of application):

* `ip.filter`: removes campaigns of "greedy" IP addresses.
* `triplet.filter`: reduces triplets to a single observation.
* `small.filter`: removes entries smaller than 1,000 bytes.
* `sequence.filter`: removes blocs of past versions.
* `size.filter`: removes entries smaller than a package's binary or source file.

These filters are off by default (e.g., ip.filter = FALSE). To apply them, set the argument for the filter you want to TRUE:

```{r,  small_filter, eval = FALSE}
packageRank(package = "cholera", small.filter = TRUE)
```

Alternatively, you can simply set `all.filters = TRUE`.

```{r,  all_filters, eval = FALSE}
packageRank(package = "cholera", all.filters = TRUE)
```

Note that the `all.filters` argument is contextual. This is because there are two sets of filters: CRAN specific functions, accessible via the `ip.filter` and `size.filter` arguments, work independently of packages, at the level of the entire log; package specific functions, accessible via the `triplet.filter`, `sequence.filter`, and `size.filter` arguments, rely on specific information about packages (e.g., size of source or binary file).

Ideally, we'd like to use both sets. However, the package specific set can be computationally expensive, especially when making relative comparisons like computing rank percentiles. This is because we need to apply the package specific filters to all the observed packages in a log, which can involve tens of thousands of packages. While not unfeasible, this will currently take a long time.

For this reason, when setting `all.filters = TRUE`, certain functions default to use only CRAN specific filters: `packageRank()`, `ipPackage()`, `countryPackage()`, `countryDistribution()` and `packageDistribution()`. Other functions default to using both CRAN and package specific functions: `packageLog()`, `packageCountry()`, and `filteredDownloads()`.

### IV - notes

#### country codes (top level domains)

While IP addresses are anonymized, `packageCountry()` and `countryPackage()` make use of the fact that the logs attempt to provide corresponding ISO country codes or top level domains (e.g., AT, JP, US). Note however, that this covers about 85% of observations (i.e., approximately 15% country codes are NA). Also, for what it's worth, there seems to be a a couple of typos for country codes: "A1" (A + number one) and "A2" (A + number 2). According to [RStudio's documentation](http://cran-logs.rstudio.com/), this coding was done using MaxMind's free database, which no longer seems to be available.

#### memoization

To avoid the bottleneck of downloading multiple log files, `packageRank()` is currently limited to individual calendar dates. To reduce the bottleneck of re-downloading logs, which can be upwards of 50 MB, ['packageRank'](https://CRAN.R-project.org/package=packageRank) makes use of memoization via the ['memoise'](https://CRAN.R-project.org/package=memoise) package.

Here's relevant code:

```{r memoization, eval = FALSE}
fetchLog <- function(url) data.table::fread(url)

mfetchLog <- memoise::memoise(fetchLog)

if (RCurl::url.exists(url)) {
  cran_log <- mfetchLog(url)
}

# Note that data.table::fread() relies on R.utils::decompressFile().
```

This means that logs are intelligently cached; those that have already been downloaded, in your current R session, will not be downloaded again.

#### time zones

The calendar date (e.g. "2021-01-01") is the unit of observation for ['packageRank'](https://CRAN.R-project.org/package=packageRank) functions. However, because the typical use case involves the _latest_ log file, time zone differences can come into play.

Let's say that it's 09:01 on 01 January 2021 and you want to compute the rank percentile for ['ergm'](https://CRAN.R-project.org/package=ergm) for the last day of 2020. You might be tempted to use the following:

```{r timezone, eval = FALSE}
packageRank(packages = "ergm")
```

However, depending on _where_ you make this request, you may not get the data you expect: in Honolulu, USA, you will; in Sydney, Australia, you won't. The reason is that you've somehow forgotten a key piece of trivia: RStudio typically posts yesterday's log around 17:00 UTC the following day.

The expression works in Honolulu because 09:01 HST on 01 January 2021 is 19:01 UTC 01 January 2021. So the log you want has been available for 2 hours. The expression fails in Sydney because 09:01 AEDT on 01 January 2021 is 31 December 2020 22:00 UTC. The log you want won't actually be available for another 19 hours.

To make life a little easier, ['packageRank'](https://CRAN.R-project.org/package=packageRank) does two things. First, when the log for the date you want is not available (due to time zone rather than server issues), you'll just get the last available log. If you specified a date in the future, you'll either get an error message or a warning that provides an estimate of when that log should be available.

Using the Sydney example and the expression above, you'd get the results for 30 December 2020:

```{r sydney, eval = FALSE}
packageRank(packages = "ergm")
```
```{r sydney_code, echo = FALSE}
packageRank(packages = "ergm", date = "2020-12-30")
```

If you had specified the date, you'd get an additional warning:

```{r sydneyB, eval = FALSE}
packageRank(packages = "ergm", date = "2021-01-01")
```
```{r sydney_codeB, echo = FALSE}
packageRank(packages = "ergm", date = "2020-12-30")
```
```
Warning message:
2020-12-31 log arrives in appox. 19 hours at 02 Jan 04:00 AEDT. Using last available!
```

Second, to help you check/remember when logs are posted in your location, there's `logDate()` and `logPostInfo()`. The former silently returns the date of the current available log. The latter adds the approximate local and UTC times when logs of the desired date are posted to RStudio's server.

Here's what you'd see using the Honolulu example:

```{r logDate, eval = FALSE}
logDate()
```
```
> [1] "2021-01-01
````

and 

```{r logPostInfo, eval = FALSE}
logPostInfo()
```

```{r logPostInfo_code, echo = FALSE, eval = FALSE}
logPostInfo(tz = "Pacific/Honolulu")
```

```
> $log.date
> [1] "2021-01-01"
>
> $GMT
> [1] "2021-01-01 17:00:00 GMT"
>
> $local
> [1] "2021-01-01 07:00:00 HST"
```

For both functions, the default is to use your time zone. To see the results in a different time zone, pass the desired zone name from `OlsonNames()` to the `tz` argument. Here are the results for Sydney when the functions are called from Honolulu (19:01 UTC):


```{r logDate2, eval = FALSE}
logDate(tz = "Australia/Sydney")
```
```
> [1] "2021-01-01"
````

and

```{r logPostInfo2, eval = FALSE}
logPostInfo(tz = "Australia/Sydney")
```

```
> $log.date
> [1] "2021-01-01"
>
> $GMT
> [1] "2021-01-01 17:00:00 GMT"
>
> $local
> [1] "2021-01-01 04:00:00 AEDT"
```

This functionality depends on R's ability to to compute your local time and time zone (e.g., `Sys.time()`). My understanding is that there may be operating system or platform specific issues that could undermine this.

#### timeout

With R 4.0.3, the timeout value for internet connections became more explicit. Here are the relevant details from that release's ["New features"](https://cran.r-project.org/doc/manuals/r-release/NEWS.html):

```
The default value for options("timeout") can be set from environment variable
R_DEFAULT_INTERNET_TIMEOUT, still defaulting to 60 (seconds) if that is not set
or invalid.
```

This change occasionally affected functions that download logs. This was especially true over slower internet connections and with larger log files. To fix this, functions that use `fetchCranLog()` will, if needed, temporarily set the timeout to 300 seconds.
