---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">",
  fig.path = "man/figures/README-"
)
```

```{r, echo = FALSE}
library(packageRank)
```

## ‘packageRank’

The '[cranlogs](https://cran.r-project.org/package=cranlogs)' package computes the number of downloads logged on RStudio's [CRAN mirror](http://cran-logs.rstudio.com). For example, we can see that the '[HistData](https://cran.r-project.org/package=HistData)' pacakge was downloaded 51 times on the first day of 2019 and 787 times in the first week:

```{r cranlogs1}
cranlogs::cran_downloads(packages = "HistData", from = "2019-01-01", to = "2019-01-01")
```

```{r cranlogs2}
cranlogs::cran_downloads(packages = "HistData", from = "2019-01-01", to = "2019-01-07")
```

In both instances, the familiar "compared to what?" question lurks in the background. Is the 51 downloads large or small? Is the pattern in the first week of January typical or unusual? The aim of 'packageRank' is to help answer these questions by providing some numerical and visual perspective.

### computation

‘packageRank’ computes a package's rank percentile ("percentile") and nominal rank ("rank"):

```{r packageRank1}
packageRank(package = "HistData", date = "2019-01-01")
```
In this case, those 51 downloads put 'HistData' in the 93rd percentile
(93% of packages had fewer downloads) and in 920th place of the 14,020 packages downloaded.^[Note that because packages with zero downloads are not recorded in the log, there is a censoring problem.]

However, the 920th place ranking is only "nominal". This is because it's possible, especially when the number of downloads is small, that mulitiple packages will have the same number of downloads. As a result, a package's rank will sometimes be determined by the fact that packages with the same number of downloads are sorted alphabetically: 'HistData''s nominal rank benefits from the fact that it appears second in the vector of packages with 51 downloads:

```{r nominal}
pkg.rank <- packageRank(package = "HistData", date = "2019-01-01")
downloads <- pkg.rank$crosstab

downloads[downloads == 51]
```

### visualization (cross-sectional)

‘packageRank’ visualizes a package's position in the distribution of a given day's downloads.

```{r plot1, fig.width = 5, fig.height = 5, fig.align = "left"}
plot(packageRank(package = "HistData", date = "2019-05-01"), graphics_pkg = "base")
```

The cross-sectional view above plots a package's rank (x-axis) against the logarithm of its downloads (y-axis). The plot highlights the selected package's relative position in the distribution along with its percentile and its number of downloads (in red). The plot also shows the location of the 75th, 50th and 25th percentiles (dotted gray vertical lines) as well as the package with the most downloads ('devtools') and the total number of downloads (2,982,767) from the CRAN mirror (both in blue).

Just like cranlogs::cran_downloads(), you can also pass a vector of packages:

```{r plot2, fig.width = 6, fig.height = 6, fig.align = "left"}
plot(packageRank(package = c("cholera", "HistData", "regtools"), date = "2019-05-01"),
  graphics_pkg = "ggplot2")
```

### visualization (longitudinal)

In more limited fashion, ‘packageRank’ visualizes a package's position in the distribution of downloads over time (currently only "last-week" and "last-month").

```{r plot_ts, fig.width = 5, fig.height = 5, fig.align = "left"}
plot(packageRankTime(package = "HistData", when = "last-month"), graphics_pkg = "base")
```
The longitudinal view above plots the date (x-axis) against the logarithm of a package's downloads (y-axis) (in red). In the background, the function plots the same data (in gray) using a cohort defined by a stratified random sample of packages.^[Within each 5% rank percentile bin (e.g., 0 to 5, 5 to 10, etc.), a random sample of 5% of packages is selected and then tracked over time as a cohort.] This cohort is used to estimate of the overall temporal pattern of package downloads.

As above, you can pass a vector of packages:

```{r plot_ts2, fig.width = 6, fig.height = 6, fig.align = "left"}
plot(packageRankTime(package = c("Rcpp", "HistData", "rlang"), when = "last-month"))
```

### memoization

To reduce the need to re-download logs for each function call, packageRank() and packageRankTime() use the 'memoise' package:

```{r memoization, eval = FALSE}
fetchLog <- function(x) data.table::fread(x)

mfetchLog <- memoise::memoise(fetchLog)

if (RCurl::url.exists(url)) {
  cran_log <- mfetchLog(url)
}
```

### plot.cranlogs()

The cran_downloads2() function essentially duplicates cranlogs::cran_downloads() but adds a class to allow for the use of S3 generic plot methods.

```{r cranlogsA, fig.width = 5, fig.height = 5, fig.align = "left"}
plot(cran_downloads2(package = c("Rcpp", "rlang", "dplyr")), graphics_pkg = "base")
```

```{r cranlogsB, fig.width = 6, fig.height = 6, fig.align = "left"}
plot(cran_downloads2(package = c("Rcpp", "rlang", "dplyr"), from = "2019-01-01", to = "2019-01-31"))
```

### base R graphics and 'ggplot2'

All plot functions are available in both base R graphics and 'ggplot2' flavors via the graphics_pkg argument ("base" or "ggplot2").


### Installation

To install the development version of 'packageRank' from GitHub:

```{r, eval = FALSE}
devtools::install_github("lindbrook/packageRank")
```

'packageRank' relies on an active internet connection.

### Notes
