---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">",
  fig.path = "man/figures/README-"
)
```

```{r library, echo = FALSE}
library(packageRank)
```

## packageRank: compute and visualize package download counts and rank percentiles

['packageRank'](https://CRAN.R-project.org/package=packageRank) is an R package that helps put package download counts into context.

It does so via two functions, `cranDownloads()` and `packageRank()`, and a set of filters that clean log files.

`cranDownloads()` extends the functionality of [`cranlogs::cran_downloads()`](http://r-hub.github.io/cranlogs/) by adding a more user-friendly interface and providing a generic R `plot()` method to makes visualization easy.

`packageRank()` uses rank percentiles, a nonparametric statistic that tells you the percentage of observations (i.e., packages) with fewer counts (i.e., downloads), to help you see how your package is doing relative to _all_ other [CRAN](https://CRAN.R-project.org/) packages.

The filters try to remove software artifacts (e.g., downloads that are too "small") and behavioral artifacts (e.g., efforts to download all of [CRAN](https://CRAN.R-project.org/)) that inflate the package download count

Note that [‘packageRank’](https://CRAN.R-project.org/package=packageRank) requires an active internet connection, and relies on the [‘cranlogs’](https://CRAN.R-project.org/package=cranlogs) package and [RStudio’s download logs](http://cran-logs.rstudio.com/). The latter record traffic to what was previously called RStudio’s [CRAN mirror](http://cran.rstudio.com) and which is now listed as the [“0-Cloud” mirror](http://cloud.r-project.org) “sponsored by RStudio”. Logs for the previous day are generally posted by 17:00 UTC. Updated results for functions that rely on ‘cranlogs’ are generally available soon thereafter.

### I - getting started

To install 'packageRank' from CRAN:

```{r cran_install, eval = FALSE}
install.packages("packageRank")
```

To install the development version from GitHub:

```{r gh_install, eval = FALSE}
# You may need to first install 'remotes' via install.packages("remotes").
remotes::install_github("lindbrook/packageRank", build_vignettes = TRUE)
```

### II - computing package download counts

`cranDownloads()` uses all the same arguments as `cranlogs::cran_downloads()`:

```{r cran_downloads, eval = FALSE}
cranlogs::cran_downloads(packages = "HistData")
```
```{r cran_downloads_code, echo = FALSE}
cranlogs::cran_downloads(packages = "HistData", from = "2020-05-01", to = "2020-05-01")
```

The only difference is that `cranDownloads()` adds four features:

#### i) "spell check" for package names

```{r spell_check_fail, eval = FALSE}
cranDownloads(packages = "GGplot2")
```
```
## Error in cranDownloads(packages = "GGplot2") :
##   GGplot2: misspelled or not on CRAN.
```

<br/>
```{r spell_check_pass, eval = FALSE}
cranDownloads(packages = "ggplot2")
```
```{r spell_check_pass_code, echo = FALSE}
cranDownloads(packages = "ggplot2", from = "2020-05-01", to = "2020-05-01")
```

<br/>
This also works for inactive or "retired" packages in the [Archive](https://CRAN.R-project.org/src/contrib/Archive):

```{r check_archive_fail, eval = FALSE}
cranDownloads(packages = "vr")
```
```
## Error in cranDownloads(packages = "vr") :
##  vr: misspelled or not on CRAN/Archive.
```

<br/>
```{r check_archive_pass, eval = FALSE}
cranDownloads(packages = "VR")
```
```{r check_archive_pass_code, echo = FALSE}
cranDownloads(packages = "VR", from = "2020-05-01", to = "2020-05-01")
```

<br/>

#### ii) two additional date formats

With `cranlogs::cran_downloads()`, you specify a time frame using the `from` and `to` arguments. The downside of this is that you _must_ use the "yyyy-mm-dd" date format. For convenience's sake, `cranDownloads()` also allows you to use "yyyy-mm" or "yyyy" (yyyy also works).

##### "yyyy-mm"
Let's say you want the download counts for ['HistData'](https://CRAN.R-project.org/package=HistData) for the month of February 2020. With `cranlogs::cran_downloads()`, you'd have to type out the whole date and remember that 2020 was a leap year:

```{r yyyy-mm_1, eval = FALSE}
cranlogs::cran_downloads(packages = "HistData", from = "2020-02-01",
  to = "2020-02-29")
```

<br/>
With `cranDownloads()`, you can just specify the year and month:

```{r yyyy-mm_2, eval = FALSE}
cranDownloads(packages = "HistData", from = "2020-02", to = "2020-02")
```

##### "yyyy" or yyyy
Let's say you want the year-to-date download counts for ['rstan'](https://CRAN.R-project.org/package=rstan). With `cranlogs::cran_downloads()`, you'd type something like:

```{r yyyy_1, eval = FALSE}
cranlogs::cran_downloads(packages = "rstan", from = "2021-01-01",
  to = Sys.Date() - 1)
```

<br/>
With `cranDownloads()`, you can use:

```{r yyyy_2, eval = FALSE}
cranDownloads(packages = "rstan", from = "2021")
```

or

```{r yyyy_3, eval = FALSE}
cranDownloads(packages = "rstan", from = 2021)
```

<br/>

#### iii) check date validity

```{r check_date, eval = FALSE}
cranDownloads(packages = "HistData", from = "2019-01-15",
  to = "2019-01-35")
```
```
## Error in resolveDate(to, type = "to") : Not a valid date.
```

#### iv) cumulative count

```{r cranDownloads_cumulative, eval = FALSE}
cranDownloads(packages = "HistData", when = "last-week")
```

```{r cranDownloads_cumulative_code, echo = FALSE}
cranDownloads(packages = "HistData", from = "2020-05-01", to = "2020-05-07")
```


<br/>

### III - visualizing package downloads

`cranDownloads()` makes visualizing package downloads easy. Just use `plot()`:

```{r cranDownloads_viz1, fig.align = "center"}
plot(cranDownloads(packages = "HistData", from = "2019", to = "2019"))
```

If you pass a vector of package names for a single day, `plot()` will return a dotchart:

```{r cranDownloads_viz2a, fig.align = "center"}
plot(cranDownloads(packages = c("ggplot2", "data.table", "Rcpp"),
  from = "2020-03-01", to = "2020-03-01"))
```

If you pass a vector of package names for multiple days, `plot()` uses `ggplot2` facets:

```{r cranDownloads_viz2, fig.align = "center"}
plot(cranDownloads(packages = c("ggplot2", "data.table", "Rcpp"),
  from = "2020", to = "2020-03-20"))
```
<br/>

If you want to plot those data in a single frame, set `multi.plot = TRUE`:

```{r cranDownloads_viz3}
plot(cranDownloads(packages = c("ggplot2", "data.table", "Rcpp"),
  from = "2020", to = "2020-03-20"), multi.plot = TRUE)
```

<br/>
If you want plot those data in separate plots on the same scale, use `graphics = "base"` and you'll be prompted for each plot:

```{r cranDownloads_viz4, eval = FALSE}
plot(cranDownloads(packages = c("ggplot2", "data.table", "Rcpp"),
  from = "2020", to = "2020-03-20"), graphics = "base")
```

If you want do the above on spearate independent scales, set `same.xy = FALSE`:

```{r cranDownloads_viz5, eval = FALSE}
plot(cranDownloads(packages = c("ggplot2", "data.table", "Rcpp"),
  from = "2020", to = "2020-03-20"), graphics = "base", same.xy = FALSE)
```

#### `packages = NULL`

`cranlogs::cran_download(packages = NULL)` computes the total number of package downloads from CRAN. You can plot these data by using:

```{r null_packages}
plot(cranDownloads(from = 2019, to = 2019))
```

#### `packages = "R"`

`cranlogs::cran_download(packages = "R")` computes the total number of downloads of the R application (note that you can only use "R" or a vector of packages names, not both!). You can plot these data by using:

```{r r_downloads}
plot(cranDownloads(packages = "R", from = 2019, to = 2019))
```

#### smoothers and confidence intervals

To add a lowess smoother to your plot, use `smooth = TRUE`:

```{r lowess}
plot(cranDownloads(packages = "rstan", from = "2019", to = "2019"),
  smooth = TRUE)
```

With graphs that use 'ggplot2', `se = TRUE` will add confidence intervals:

```{r ci}
plot(cranDownloads(packages = c("HistData", "rnaturalearth", "Zelig"),
  from = "2020", to = "2020-03-20"), smooth = TRUE, se = TRUE)
```

#### package and R release dates

To annotate a graph with a package's release dates:

```{r pkg_release_date}
plot(cranDownloads(packages = "rstan", from = "2019", to = "2019"),
  package.version = TRUE)
```

To annotate a graph with R release dates:

```{r r_release_date}
plot(cranDownloads(packages = "rstan", from = "2019", to = "2019"),
  r.version = TRUE)
```

#### plot growth curves (cumulative download counts)

To plot growth curves, set `statistic = "cumulative"`:

```{r cranDownloads_growth_curves}
plot(cranDownloads(packages = c("ggplot2", "data.table", "Rcpp"),
  from = "2020", to = "2020-03-20"), statistic = "cumulative",
  multi.plot = TRUE, points = FALSE)
```

#### population plot

To visualize a package's downloads relative to "all" other packages over time:

```{r pop_plot, eval = FALSE}
plot(cranDownloads(packages = "HistData", from = "2020", to = "2020-03-20"),
  population.plot = TRUE)
```

```{r pop_plot_code, echo = FALSE}
plot(cranDownloads(packages = "HistData", from = "2020", to = "2020-03-20"),
  population.plot = TRUE, population.seed = 1)
```

This longitudinal view of package downloads plots the date (x-axis) against the logarithm of a package's downloads (y-axis). In the background, the same variable are plotted (in gray) using a stratified random sample of packages: within each 5% interval of rank percentiles (e.g., 0 to 5, 5 to 10, 95 to 100, etc.), a random sample of 5% of packages is selected and tracked. This graphically approximates the "typical" pattern of downloads on CRAN for the selected time period.

### IV - computing package download rank percentiles

Looking at nominal download count data leads one to the "compared to what?" question. For instance, consider the data for the first week of March 2020:

```{r motivation, eval = FALSE}
plot(cranDownloads(packages = "cholera", from = "2020-03-01",
  to = "2020-03-07"))
```

```{r motivation_code, echo = FALSE, fig.align = "center"}
par(mar = c(5, 4, 4, 4))
plot(cranDownloads(packages = "cholera", from = "2020-03-01",
  to = "2020-03-07"))
par(mar = c(5, 4, 4, 2))
```

Do Wednesday and Saturday reflect surges of interest in the package or surges of traffic to [CRAN](https://CRAN.R-project.org/)? To put it differently, how can we know if a given download count is typical or unusual? One way to answer these questions is to locate your package in the frequency distribution of download counts.

Below are the distributions of logarithm of download counts for Wednesday and Saturday. The location of a vertical segment along the x-axis represents a download count and the height of a segment represents that download count's frequency. The location of ['cholera'](https://CRAN.R-project.org/package=cholera) in the distribution is highlighted in red.

```{r packageDistribution, echo = FALSE}
plot_package_distribution <- function(dat, xlim, ylim) {
  freq.dist <- dat$freq.dist
  freqtab <- dat$freqtab
  plot(freq.dist$count, freq.dist$frequency, type = "h", log = "x",
    xlab = "Downloads", ylab = "Frequency", xlim = xlim, ylim = ylim)
  axis(3, at = freqtab[1], cex.axis = 0.8, padj = 0.9, col.axis = "dodgerblue",
    col.ticks = "dodgerblue", labels = paste(names(freqtab[1]), "=",
    format(freqtab[1], big.mark = ",")))
  abline(v = freqtab[1], col = "dodgerblue", lty = "dotted")

  if (!is.null(dat$package)) {
    pkg.ct <- freqtab[names(freqtab) == dat$package]
    pkg.bin <- freqtab[freqtab == pkg.ct]
    axis(3, at = pkg.ct, labels = format(pkg.ct, big.mark = ","),
      cex.axis = 0.8, padj = 0.9, col.axis = "red", col.ticks = "red")
    abline(v = pkg.ct, col = grDevices::adjustcolor("red", alpha.f = 0.5))
    day <- weekdays(as.Date(dat$date), abbreviate = TRUE)
    title(paste0(dat$package, " @ ", dat$date, " (", day, ")"))
  } else title(paste("Distribution of Package Download Counts:", dat$date))
}

distn.data <- lapply(c("2020-03-04", "2020-03-07"), function(date) packageDistribution(package = "cholera", date = date))

xlim <- range(lapply(distn.data, function(x) x$freq.dist$count))
ylim <- range(lapply(distn.data, function(x) x$freq.dist$frequency))
```

```{r packageDistribution_wed, eval = FALSE}
plot(packageDistribution(package = "cholera", date = "2020-03-04"))
```
```{r packageDistribution_wed_code, echo = FALSE, fig.align = "center"}
plot_package_distribution(distn.data[[1]], xlim, ylim)
```

```{r packageDistribution_sat, eval = FALSE}
plot(packageDistribution(package = "cholera", date = "2020-03-07"))
```
```{r packageDistribution_sat_code, echo = FALSE, fig.align = "center"}
plot_package_distribution(distn.data[[2]], xlim, ylim)
```

While these plots give us a better picture of where ['cholera'](https://CRAN.R-project.org/package=cholera) is located, comparisons between Wednesday and Saturday are impressionistic at best: all we can confidently say is that the download counts for both days were greater than the mode.

To facilitate interpretation and comparison, I use the _rank percentile_ of a download count in place of the nominal download count. This nonparametric statistic tells you the percentage of packages with fewer downloads. In other words, it gives you the location of your package relative to the locations of all other packages. More importantly, by rescaling download counts to lie on the bounded interval between 0 and 100, rank percentiles make it easier to compare packages within and across distributions.

For example, we can compare Wednesday ("2020-03-04") to Saturday ("2020-03-07"):

```{r packageRank1}
packageRank(package = "cholera", date = "2020-03-04", ip.filter = FALSE, small.filter = FALSE)
```

On Wednesday, we can see that ['cholera'](https://CRAN.R-project.org/package=cholera) had 38 downloads, came in 5,556th place out of 18,038 unique packages downloaded, and earned a spot in the 68th percentile.

```{r packageRank2}
packageRank(package = "cholera", date = "2020-03-07", ip.filter = FALSE, small.filter = FALSE)
```

On Saturday, we can see that ['cholera'](https://CRAN.R-project.org/package=cholera) had 29 downloads, came in 3,061st place out of 15,950 unique packages downloaded, earned a spot in the 80th percentile.

So contrary to what the nominal counts tell us, one could say that the interest in ['cholera'](https://CRAN.R-project.org/package=cholera) was actually greater on Saturday than on Wednesday.

#### computing rank percentile

To compute rank percentiles, I do the following. For each package, I tabulate the number of downloads and then compute the percentage of packages with fewer downloads. Here are the details using ['cholera'](https://CRAN.R-project.org/package=cholera) from Wednesday as an example:

```{r percentile}
pkg.rank <- packageRank(packages = "cholera", date = "2020-03-04",
  ip.filter = FALSE, small.filter = FALSE)

downloads <- pkg.rank$freqtab

round(100 * mean(downloads < downloads["cholera"]), 1)
```

To put it differently:

```{r percentile2}
(pkgs.with.fewer.downloads <- sum(downloads < downloads["cholera"]))

(tot.pkgs <- length(downloads))

round(100 * pkgs.with.fewer.downloads / tot.pkgs, 1)
```

#### nominal ranks

In the example above, 38 downloads puts 'cholera' in 5,556th place among the 18,038 packages downloaded. This rank is "nominal" because it's possible that multiple packages can have the same number of downloads. As a result, a package's nominal rank (but not its rank percentile) can be affected by its name: packages with the same number of downloads are sorted in alphabetical order. Thus, 'cholera' benefits from the fact that it is 31st in the list of 263 packages with 38 downloads:

```{r nominal}
pkg.rank <- packageRank(packages = "cholera", date = "2020-03-04",
  ip.filter = FALSE, small.filter = FALSE)
downloads <- pkg.rank$freqtab

which(names(downloads[downloads == 38]) == "cholera")
length(downloads[downloads == 38])
```

### V - visualizing package download rank percentiles

To visualize `packageRank()`, use `plot()`.

```{r packageRank_plot_wed, eval = FALSE}
plot(packageRank(packages = "cholera", date = "2020-03-04"))
```

```{r packageRank_data, echo = FALSE}
dat <- lapply(c("2020-03-04", "2020-03-07"), function(x) {
  packageRank("cholera", date = x, ip.filter = FALSE, small.filter = FALSE)
})

freqtab1 <- dat[[1]]$freqtab
freqtab2 <- dat[[2]]$freqtab
xlim <- range(seq_along(freqtab1), seq_along(freqtab2))
ylim <- range(c(freqtab1), c(freqtab2))
```

```{r packageRank_plot_code_wed, echo = FALSE, fig.align = "center"}

freqtab <- dat[[1]]$freqtab
package.data <- dat[[1]]$package.data
pkg <- dat[[1]]$packages
date <- dat[[1]]$date
y.max <- freqtab[1]
q <- stats::quantile(freqtab)[2:4]

iqr <- vapply(c("75%", "50%", "25%"), function(id) {
  dat <- which(freqtab > q[[id]])
  dat[length(dat)]
}, numeric(1L))

plot(seq_along(freqtab), c(freqtab), type = "l", xlab = "Rank",
  ylab = "log10(Count)", log = "y", xlim = xlim, ylim = ylim)
abline(v = iqr, col = "black", lty = "dotted")
iqr.labels <- c("75th", "50th", "25th")
invisible(lapply(seq_along(iqr), function(i) {
  text(iqr[[i]], y.max / 2, labels = iqr.labels[i], cex = 0.75)
}))
abline(v = which(names(freqtab) == pkg), col = "red")
abline(h = freqtab[pkg], col = "red")
pct <- package.data[package.data$package == pkg, "percentile"]
pct.label <- paste0(round(pct, 2), "%")
axis(3, at = which(names(freqtab) == pkg), padj = 0.9, col.axis = "red",
  col.ticks = "red", labels = pct.label, cex.axis = 0.8)
axis(4, at = freqtab[pkg], col.axis = "red", col.ticks = "red",
  cex.axis = 0.8, labels = format(freqtab[pkg], big.mark = ","))
points(which(names(freqtab) == pkg), freqtab[pkg], col = "red")
points(which(names(freqtab) == names(freqtab[1])), y.max,
  col = "dodgerblue")
text(which(names(freqtab) == names(freqtab[1])), y.max, pos = 4,
  labels = paste(names(freqtab[1]), "=", format(freqtab[1],
  big.mark = ",")), cex = 0.8, col = "dodgerblue")
text(max(xlim), max(ylim),
  labels = paste("Tot = ", format(sum(freqtab), big.mark = ",")), cex = 0.8,
  col = "dodgerblue", pos = 2)
day <- weekdays(as.Date(date), abbreviate = TRUE)
title(main = paste0(pkg, " @ ", date, " (", day, ")"))
```

<br/>

```{r packageRank_plot_sat, eval = FALSE}
plot(packageRank(packages = "cholera", date = "2020-03-07"))
```

```{r packageRank_plot_code_sat, echo = FALSE, fig.align = "center"}

freqtab <- dat[[2]]$freqtab
package.data <- dat[[2]]$package.data
pkg <- dat[[2]]$packages
date <- dat[[2]]$date
y.max <- freqtab[1]
q <- stats::quantile(freqtab)[2:4]

iqr <- vapply(c("75%", "50%", "25%"), function(id) {
  dat <- which(freqtab > q[[id]])
  dat[length(dat)]
}, numeric(1L))

plot(seq_along(freqtab), c(freqtab), type = "l", xlab = "Rank",
  ylab = "log10(Count)", log = "y", xlim = xlim, ylim = ylim)
abline(v = iqr, col = "black", lty = "dotted")
iqr.labels <- c("75th", "50th", "25th")
invisible(lapply(seq_along(iqr), function(i) {
  text(iqr[[i]], y.max / 2, labels = iqr.labels[i], cex = 0.75)
}))
abline(v = which(names(freqtab) == pkg), col = "red")
abline(h = freqtab[pkg], col = "red")
pct <- package.data[package.data$package == pkg, "percentile"]
pct.label <- paste0(round(pct, 2), "%")
axis(3, at = which(names(freqtab) == pkg), padj = 0.9, col.axis = "red",
  col.ticks = "red", labels = pct.label, cex.axis = 0.8)
axis(4, at = freqtab[pkg], col.axis = "red", col.ticks = "red",
  cex.axis = 0.8, labels = format(freqtab[pkg], big.mark = ","))
points(which(names(freqtab) == pkg), freqtab[pkg], col = "red")
points(which(names(freqtab) == names(freqtab[1])), y.max,
  col = "dodgerblue")
text(which(names(freqtab) == names(freqtab[1])), y.max, pos = 4,
  labels = paste(names(freqtab[1]), "=", format(freqtab[1],
  big.mark = ",")), cex = 0.8, col = "dodgerblue")
text(max(xlim), max(ylim),
  labels = paste("Tot = ", format(sum(freqtab), big.mark = ",")), cex = 0.8,
  col = "dodgerblue", pos = 2)
day <- weekdays(as.Date(date), abbreviate = TRUE)
title(main = paste0(pkg, " @ ", date, " (", day, ")"))
```

These graphs, customized to be on the same scale, plot the _rank order_ of packages' download counts (x-axis) against the logarithm of those counts (y-axis). It then highlights a package's position in the distribution along with its rank percentile and download count (in red). In the background, the 75th, 50th and 25th percentiles are plotted as dotted vertical lines. The package with the most downloads, ['magrittr'](https://CRAN.R-project.org/package=magrittr) in both cases, is at top left (in blue). The total number of downloads is at the top right (in blue).


### VII - miscellanea

#### country codes (top level domains)

While IP addresses are anonymized, the logs do attempt to provide the corresponding ISO country codes or top level domains (e.g., AT, JP, US). This covers about 85% of observations (i.e., approximately 15% country codes are NA). Also, for what it's worth, there seems to be a a couple of typos for country codes: "A1" (A + number one) and "A2" (A + number 2). According to [RStudio's documentation](http://cran-logs.rstudio.com/), this coding was done using MaxMind's free database, which no longer seems to be available.

#### memoization

To avoid the bottleneck of downloading multiple log files, `packageRank()` is currently limited to individual calendar dates. To reduce the bottleneck of re-downloading logs, which can be upwards of 50 MB, 'packageRank' makes use of memoization via the 'memoise' package.

Here's relevant code:

```{r memoization, eval = FALSE}
fetchLog <- function(url) data.table::fread(url)

mfetchLog <- memoise::memoise(fetchLog)

if (RCurl::url.exists(url)) {
  cran_log <- mfetchLog(url)
}

# Note that data.table::fread() relies on R.utils::decompressFile().
```

This means that logs are intelligently cached; those that have already been downloaded, in your current R session, will not be downloaded again.

#### time zones

The calendar date (e.g. "2021-01-01") is the natural unit of observation for ['packageRank'](https://CRAN.R-project.org/package=packageRank) functions. However, because the typical use case involves getting the most recent data (i.e., the _latest_ log), time zone differences can come into play.

Let's say that it's 09:01 on 01 January 2021 and you want to compute the rank percentile for the ['ergm'](https://CRAN.R-project.org/package=ergm) package for the last day of 2020. You might be tempted to enter the following expression:

```{r timezone, eval = FALSE}
packageRank(packages = "ergm")
```

However, depending on _where_ you make this request, you may not get the data you expect. If you're in Honolulu, USA, you will. If you're in Sydney, Australia, you won't. The reason is that you've somehow forgotten a key piece of trivia: RStudio typically posts yesterday's log by 17:00 UTC the following day.

The expression works in Honolulu because 09:01 HST on 01 January 2021 is 19:01 UTC 01 January 2021: the log you want has been available for 2 hours. The expression fails in Sydney because 09:01 AEDT on 01 January 2021 is 31 December 2020 22:00 UTC: the log you want won't actually be posted for another 19 hours.

To make life a little easier, ['packageRank'](https://CRAN.R-project.org/package=packageRank) does two things. First, when the date of the log you want is not available (due to time zone rather than server issues), you'll get the last available log as a substitute and a warning that provides an estimate of when that log should be available.

For the Honolulu example, you'd get the following:

```{r honolulu, echo = FALSE}
packageRank0(packages = "ergm", date = "2020-12-31")
```

For the Sydney example, you'll get the results for 30 December 2020 and the following message:

```{r sydney, echo = FALSE}
packageRank0(packages = "ergm", date = "2020-12-30")
```
```
Warning message:
2020-12-31 log arrives in appox. 19 hours at 02 Jan 04:00 AEDT. Using last available!
```

Second, to help you remember when logs are posted in your locale, there's `logPostInfo()`. At the time it's run, it gives you the date of the latest available log along with the local and UTC time when that log should b posted to RStudio's server.

Here's what you'd see in Honoloulu:

```{r logPostInfo, eval = FALSE}
logPostInfo()
```

```{r logPostInfo_code, echo = FALSE, eval = FALSE}
logPostInfo(tz = "Pacific/Honolulu")
```

```
> $log.date
> [1] "2021-02-01"
>
> $GMT
> [1] "2021-02-02 17:00:00 GMT"
>
> $local
> [1] "2021-02-02 07:00:00 HST"
```

The default is to use your local time zone, via Sys.timezone(). To use a specific time zone, pass the desired zone name from the list in OlsonNames() to the `tz` argument:

```{r logPostInfo2, eval = FALSE}
logPostInfo(tz = "Australia/Sydney")
```

```
> $log.date
> [1] "2021-02-01"
>
> $GMT
> [1] "2021-02-02 17:00:00 GMT"
>
> $local
> [1] "2021-02-03 04:00:00 AEDT"
```

This functionality depends on R's ability to to compute your clock time and time zone (e.g., Sys.time()). My understanding is that there may be operating system or platform related issues that could affect this.

#### timeout

With R 4.0.3, the timeout value for internet connections became more explicit. Here's the revelant details from this release's ["New features"](https://cran.r-project.org/doc/manuals/r-release/NEWS.html):

```
The default value for options("timeout") can be set from enviromnent variable
R_DEFAULT_INTERNET_TIMEOUT, still defaulting to 60 (seconds) if that is not set
or invalid.
```

This change occasionally affected functions that download logs. This was especially true over slower internet connections and with larger log files. To fix this, relevant functions, which make use `fetchCranLog()`, will now temporarily change the timeout to 300 seconds and then reset the value back to your configuration's original setting.
