---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">",
  fig.path = "man/figures/README-"
)
```

```{r, echo = FALSE}
library(packageRank)
```

## packageRank: compute and visualize package download counts and percentiles

### features

* compute and visualize the counts and ranks (nominal and percentile) of downloads from RStudio's [CRAN mirror](http://cran-logs.rstudio.com) or from [Bioconductor](https://bioconductor.org/).
* provide S3 plot method sfor `cranlogs::cran_downloads()`.

NOTE: 'packageRank' relies on an active internet connection.

### getting started

To install 'packageRank' from CRAN:

```{r, eval = FALSE}
install.packages("packageRank")
```

To install the latest development version from GitHub:

```{r, eval = FALSE}
# You may need to first install the 'remotes' via install.packages("remotes").

remotes::install_github("lindbrook/packageRank", build_vignettes = TRUE)
```

### background

The '[cranlogs](https://cran.r-project.org/package=cranlogs)' package computes the raw number of downloads of packages or R using RStudio's [CRAN mirror](http://cran-logs.rstudio.com).

For example, we can see that on the first day of 2019 the '[HistData](https://cran.r-project.org/package=HistData)' package was downloaded 51 times:

```{r cranlogs1}
cranlogs::cran_downloads(packages = "HistData", from = "2019-01-01", to = "2019-01-01")
```

And 787 times in the first week:

```{r cranlogs2}
cranlogs::cran_downloads(packages = "HistData", from = "2019-01-01", to = "2019-01-07")
```

In both cases, the "compared to what?" question lurks in the background. Is 51 downloads large or small? Is the pattern that first week typical or unusual? To answer these questions, 'packageRank' puts package download counts into greater context.

### compute percentiles and ranks

To do so, we can compute the rank percentile and nominal rank of a package's downloads:

```{r packageRank0, echo = FALSE}
packageRank(packages = "HistData", date = "2019-01-01")
```

```{r packageRank1}
packageRank(packages = "HistData", date = "2019-01-01")
```

Here, we see that 51 downloads put 'HistData' in the 93rd percentile. This statistic, familiar to anyone who's taken a standardized test, tell us that 93% of packages had fewer downloads than 'HistData':^[Note that because packages with zero downloads are not recorded in the log, there is a censoring problem.]

```{r percentile}
pkg.rank <- packageRank(packages = "HistData", date = "2019-01-01")
downloads <- pkg.rank$crosstab

round(100 * mean(downloads < downloads["HistData"]), 1)

# OR

(pkgs.with.fewer.downloads <- sum(downloads < downloads["HistData"]))

(tot.pkgs <- length(downloads))

round(100 * pkgs.with.fewer.downloads / tot.pkgs , 1)
```

We also see that 51 downloads put 'HistData' in 920th place among the 14,020 packages with at least one download. This rank is "nominal" because multiple packages can have the same number of downloads. As a result, a package's nominal rank (but not its rank percentile) can be affected by its names: packages with the same number of downloads are sorted in alphabetical order. For the case at hand, 'HistData' benefits from the fact that it is second in the list of packages with 51 downloads:

```{r nominal}
pkg.rank <- packageRank(packages = "HistData", date = "2019-01-01")
downloads <- pkg.rank$crosstab

downloads[downloads == 51]
```

With this release, 'packageRank' offers a version of this function for Bioconductor:

```{r packageRank2}
bioconductorRank(packages = "monocle", date = "2019-01")
```

Note the difference in date format: the logs on Bioconductor are aggreagated either to the month or to the year.

### visualization (cross-sectional)

To visualize a package's relative position using ranks, use the S3 generic plot method:

```{r plot1a, fig.width = 5, fig.height = 5, fig.align = "left"}
plot(packageRank(packages = "HistData", date = "2019-01-01"))
```

```{r plot1b, fig.width = 5, fig.height = 5, fig.align = "left"}
plot(bioconductorRank(packages = "cicero", date = "2019-01"))
```

This cross-sectional view plots a package's rank (x-axis) against the base 10 logarithm of its downloads (y-axis) and highlights its position in the overall distribution of downloads. In addition, the plot illustrates 1) a package's rank percentile and raw count of downloads (in red); 2) the location of the 75th, 50th and 25th percentiles (dotted gray vertical lines); 3) the package with the most downloads, in this case 'WGCNA' for CRAN and 'BiocInstaller' for Bioconductor (in blue, top left); and 4) the total number of downloads (2,982,767 for CRAN; 1,854,793 for Bioconductor) (in blue, top right).

You can even pass a vector of packages:

```{r plot2a, fig.width = 6, fig.height = 6, fig.align = "left"}
plot(packageRank(packages = c("cholera", "HistData", "regtools"), date = "2019-01-01"))
```

```{r plot2b, fig.width = 6, fig.height = 6, fig.align = "left", eval = FALSE, echo = FALSE}
plot(bioconductorRank(packages = c("cicero", "graph", "monocle"), date = "2019-01"))
```

### visualization (longitudinal)

To visualize a package's relative position over time, use `packageRankTime()`:

```{r plot_ts, fig.width = 5, fig.height = 5, fig.align = "left", eval = TRUE}
plot(packageRankTime(packages = "HistData", when = "last-month"), graphics_pkg = "base")
```

This longitudinal view plots the date (x-axis) against the logarithm of a package's downloads (y-axis).

In the background, the same variable are plotted (in gray) for a stratified random sample of packages.^[Within each 5% interval of rank percentiles (e.g., 0 to 5, 5 to 10, 95 to 100, etc.), a random sample of 5% of packages is selected and tracked over time.] This sample approximates the "typical" pattern of package downloads for that time period.

As above, you can pass a vector of packages:

```{r plot_ts2, fig.width = 6, fig.height = 6, fig.align = "left", eval = FALSE}
plot(packageRankTime(packages = c("Rcpp", "HistData", "rlang"), when = "last-month"))
```

Note that only two time frames are available: "last-week" and "last-month". Also, this function is not currently available for Bioconductor: the CRAN version has server side support via the 'cranlogs' package.

### computing download counts

To compute or plot package download counts, 'packageRank' offers a couple of choices. If you want to use 'cranlogs' style syntax, use 'cran_downloads2()'. The only difference between this function and 'cranlogs::cran_downloads()' is that 'cran_downloads2()' has an generic S3 plot method.

```{r cran_downloads}
cranlogs::cran_downloads(packages = "HistData", when = "last-week")
```

```{r cran_downloads2}
cran_downloads2(packages = "HistData", when = "last-week")
plot(cran_downloads2(packages = "HistData", when = "last-week"))
```

Otherwise, I recommend using `cranDownloads()`. Not only does it have a generic S3 plot method, but it has a more flexible interface, especially for the 'for' and 'to' arguments. With 'cranlogs::cran_downloads', you have to write out the entire 8 character date. With 'cranDownloads()', you can provide dates as: yyyy-mm-dd, yyyy-mm or yyyy.

```{r cranDownloads}
# Downloads from January through June 2019
plot(cranDownloads(packages = "HistData", from = "2019-01-15", to = "2019-06-25"), points = FALSE, smooth = TRUE)

# Downloads from January through June 2019
plot(cranDownloads(packages = "HistData", from = "2019-01", to = "2019-06"), points = FALSE, smooth = TRUE)

# Year-to-date 
plot(cranDownloads(packages = "HistData", from = "2019"), points = FALSE, smooth = TRUE)
```

### graphics: base R and 'ggplot2'

All plot are available as both base R and 'ggplot2' graphs. By default, plot with single frame/panels (one package or one day) use base graphics while those with multiple frames/panels use 'ggplot2'. You can override these defaults by using the "graphics" argument in the plot() method.

### memoization

To avoid the bottleneck of downloading multiple log files, `packageRank()` is limited to individual days. However, to reduce the need to re-download logs for a given day, 'packageRank' makes use of memoization via the 'memoise' package.

Here's relevant code:

```{r memoization, eval = FALSE}
fetchLog <- function(x) data.table::fread(x)

mfetchLog <- memoise::memoise(fetchLog)

if (RCurl::url.exists(url)) {
  cran_log <- mfetchLog(url)
}
```

If you use `fetchLog()`, the log file, which can sometimes be as large as 50 MB, will be downloaded every time you call the function. If you use `mfetchLog()`, logs are intelligently cached; those that have already been downloaded, in your current R session, will not be downloaded again.


### Notes
