---
title: "Note on queryPercentile()"
author: "lindbrook"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">",
  fig.path = "figures/query_percentile-"
)
```

```{r library, echo = FALSE}
library(packageRank)
```

Due to the discrete nature of counts, your choice of percentile may not be available. This is because they may fall in the vertical gaps in the observed data:

```{R, discrete_counts, echo = FALSE}
tmp <- cranDistribution(date = "2024-08-01")
plot(tmp$data$percentile, pch = 46, xlab = "Nominal Rank", ylab = "Percentile Rank", main = "Observed Percentile Ranks")
```

For this reason, `queryPercentile()` rounds your selection to whole numbers. Also, be aware that the default value, which is set to 50, uses `median()`to guarantee a result.

```{R, rank_percentile, eval = FALSE}
head(queryPercentile())
```
```{R, rank_percentile_code, echo = FALSE}
# head() is used because there will be many observations with median count.
head(queryPercentile(date = "2024-08-01"))
```

You can also set a range of percentile ranks using the 'lo' and/or 'hi' arguments. If you get an error message, you may need to widen your interval:

```{R, lo_hi, eval = FALSE}
head(queryPercentile(lo = 95, hi = 96), 3)
tail(queryPercentile(lo = 95, hi = 96), 3)
```
```{R, lo_hi_code, echo = FALSE}
head(queryPercentile(lo = 95, hi = 96, date = "2024-08-01"), 3)
tail(queryPercentile(lo = 95, hi = 96, date = "2024-08-01"), 3)
```