---
title: "Downloading Error with R version 4.0.3"
author: "lindbrook"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Downloading error

R version 4.0.3 seems to create a problem that prevents the downloading of log files. Reverting back to version 4.0.2 is the working solution. 

Here's the code for replication.

## Code for URL

```{r error_replication, eval = FALSE}
date <- Sys.Date() - 1
year <- as.POSIXlt(date)$year + 1900
rstudio.url <- "http://cran-logs.rstudio.com/"
log.url <- paste0(rstudio.url, year, '/', date, ".csv.gz")
```

## fread() code to download log file 

The current implementation in 'packageRank' uses the following:

```{r fetch, eval = FALSE}
cran_log <- data.table::fread(log.url)
```

## Base R code to download log file 

The equivalent in base R is:

```{r fetch_base, eval = FALSE}
connection <- gzcon(url(log.url))
dat <- readLines(connection)
cran_log <- read.csv(textConnection(dat))
```

## Error with fread()

```
Error in download.file(input, tmpFile, method = method, mode = "wb", quiet = !showProgress) :
  download from 'http://cran-logs.rstudio.com/2020/2020-10-14.csv.gz' failed
In addition: Warning messages:
1: In download.file(input, tmpFile, method = method, mode = "wb", quiet = !showProgress) :
  downloaded length 69100268 != reported length 77777704
2: In download.file(input, tmpFile, method = method, mode = "wb", quiet = !showProgress)
```

## Error with base R

```
Error in readLines(connection) : cannot read from connection
In addition: Warning messages:
1: In for (i in seq_along(specs)) { :
  closing unused connection 3 (http://cran-logs.rstudio.com/2020/2020-10-14.csv.gz)
2: In readLines(connection) :
  URL 'http://cran-logs.rstudio.com/2020/2020-10-14.csv.gz': Timeout of 60 seconds was reached
```
